---
title: "Lab2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "C:/Users/michailfragkias/Dropbox/Course.ECON521.GradEconometrics/Fall2022_SharedWithStudents/Labs")
options(repos=c("<https://mirror.las.iastate.edu/CRAN/>"))
```

## Lab 2

```{r}
rm(list=ls())
```

## Variables, vectors, assignments of values, descriptive stats

Let's start again with vector assignments - numeric values
```{r}
tree.heights <- c(4.3,7.1,6.3,5.2,3.2,2.1)
```

Note that the period can help in the naming variables
```{r}
tree.widths <- c(24.5, 30.2, 29.5, 26.7, 23.1, 21.5)
tree.heights+tree.widths ##vector addition- does not make sense here :)
z<-c(tree.heights,tree.widths)
z
sum(tree.heights) ##summation of elements
mean(tree.heights)
sd(tree.heights)
length(tree.heights) ##length of vector
max.tree.height <- max(tree.heights)
max.tree.height
```

We can pick subsets of vectors, which can also be transformed on the fly.
```{r}
tree.heights[1]
tree.heights[1:3]
sqrt(tree.heights[1:3])
tree.heights[c(5,3,2)]
```

Let's look at vector assignments with character values or logical values
```{r}
name <- "Lex Luthor"
name
cities <- c("Seattle", "San Diego", "Portland", "Boise")
length(cities) # number of elements in the cities vector
northern <- c(TRUE, FALSE, TRUE, TRUE) #an example of a logical variable
cities[northern] #subseting using a logical variable
```

## Important Data Types
We will work with only a subset of data types that R can handle, primarily character, numeric and logical.

The character() command creates vector of the specified length with each element equal to an empty character:
```{r}
character(8) 
length(character(8))
as.character("8") #conversion of numeric to character
is.character(8) #test
is.character("8") #another test
```

The numeric() command creates a vector of the specified length with each element equal to zero

```{r}
numeric(8) 
as.numeric(c("1980", "-8", "Geography")) #conversions
as.numeric(c(FALSE, TRUE))
is.numeric(c(8,8))
is.numeric(c(8,8,8,"8"))
```

The logical() command creates vector of specified length with each element equal to FALSE
```{r}
logical(7) 
!logical(7) #negation
as.logical(c(7, 5, 0, -4, 5)) #conversion (non-zero elements are TRUE)
as.logical(c(7, 5, 0, -4, 5)) * 1 #conversion with numeric logical values (0/1)
as.logical(c(7, 5, 0, -4, 5)) + 0 #same thing
as.logical(c("True", "T", "t", "FALSE", "Raspberry","9", "0", 0)) #notice what gets recognized
```

Logical vectors can be used for indexing
```{r}
data <- c(3, 6, 9, 99, 54, 32, -102)
index <- (data>10) # create an logical index of values greater than 10 in the dataset
index

data[index]
data[!index]
sum(data)
sum(data[index])
```

## Important Data Classes
The Data Types we went over populate different data structures (or classes). We will work primarily with vectors, matrices, data frames, lists and factors

###Vectors

This is how you can alternatively create a vector with type and length defined: vector(mode="numeric", length=8). Note that if you don't define the mode, you will get logical
```{r}
vector(length=8) 
```

You can test if the class is vector. Let's create something that is not a vector
```{r}
tmp <- data.frame (a=10:15, b=15:20)
tmp
is.vector(tmp)
is.vector(data)
#as.vector(tmp)
```

###Matrices
The matrix() command creates arrays that we recognize as matrices. We will work on matrix algebra later on in the semester.
```{r}
matrix(ncol=2,nrow=2)
matrix(1:6) #matrix dimensions not defined
matrix(1:6,ncol=2) #matrix dimensions using only ncol
matrix(1:6, c(3,2)) #matrix dimensions using concatenate
as.matrix(6:3)
as.matrix(c(6:3)) #equivalent to the above
is.matrix(as.matrix(6:3))
```

Matrix rows and columns can be named
```{r}
flow <- matrix(1:9, c(3,3), byrow=TRUE) #byrow=TRUE fills the matrix by rows
flow
colnames(flow) <- c("Col1","Col2","Col3")
rownames(flow) <- c("Row1", "Row2", "Row3")
flow
```

There are several functions that can summarize row or column data
```{r}
sumsofrows <- rowSums(flow)
sumsofrows
```

But not that if the data class is not matrix, then you can name elements by the command names().
```{r}
z <- c(6,7,8)
names(z) <- c("Six", "Seven", "Eight")
z
```

## Exploring Data
We will work with an example, the Phillips dataset, available as a .dta file

If you need to, set your working directory:
setwd("/Users/michailfragkias/Dropbox/Course.ECON521.GradEconometrics/Fall2020_ECON521_SharedWithStudents/Labs")

```{r}
library(foreign)
phillips <- read.dta("Data/wool/phillips.dta")
summary(phillips) # Provides basic descriptive statistics and frequencies.
edit(phillips) # Open data editor - DON'T DO THAT!!!
str(phillips) # Provides the structure of the dataset 
names(phillips) # Lists variables in the dataset 
head(phillips) # First 6 rows of dataset
head(phillips, n=10)# First 10 rows of dataset
head(phillips, n= -10) # All rows but the last 10 
tail(phillips) # Last 6 rows
tail(phillips, n=10) # Last 10 rows
tail(phillips, n= -10) # All rows but the first 10 
phillips[1:10, ] # First 10 rows
phillips[1:10,1:3] # First 10 rows of data of the first 3 variables
plot(unem ~ inf, data=phillips)
```

## Simple Plots
```{r}
library(graphics)
x <- seq(100)
y <- x^0.5
plot(x,y)
plot(x,y, type="p") #this is the default
plot(x,y, type="l") #continuous lines
plot(x,y, type="b") #for both
plot(x,y, type="o") #for both overplotted
plot(x,y, type="s") #for stair steps

barplot(y,col="lightgreen")

x <- c(1:5); y <- x # create some data 
par(pch=22, col="red") # plotting symbol and color 
par(mfrow=c(2,4)) # all plots on one page 
opts = c("p","l","o","b","c","s","S","h") 
for(i in 1:length(opts)){ 
  heading = paste("type=",opts[i]) 
  plot(x, y, type="n", main=heading) 
  lines(x, y, type=opts[i]) 
}
```

# Creating a supply and demand plot
```{r}
Q  <- seq(0,150)
Ps <- -(10/40) + (1/40)*Q
Pd <- (100/20) - (1/20)*Q
plot(Q,Ps,xlab="Quantity (kg)",  ylab="Price (euros per kg)",type="l")
lines(Q,Pd)
text(130,2.6,"Supply")
text(120,0,"Demand")
# A simple way to add a straight line to an existing plot uses the abline function. 
# Read abline as a-b line: a is the intercept and b is the slope or gradient of the line with equation y = a + bx. 
# For instance, suppose that the demand schedule shifts to: 
abline(a = 120/20, b = -1/20, lty = "dashed")
```

## Example: Pedometer Data
```{r}
peddata <-read.table("Data/pedometer.csv", header=TRUE, sep=",")
head(peddata)
x <- hist(peddata$Steps,col="lightblue")
x <- hist(peddata$Steps,breaks=20,col="lightblue")
plot(peddata$Steps ~ peddata$Observation,col="blue")
#
aggregate(peddata$Steps~peddata$Day, FUN=mean)

myModel <- lm(peddata$Steps ~ peddata$Observation)
summary(myModel)
lines(fitted(myModel))
#lines(fitted(loess(myModel)),col="red")
```

## Anscombe Quartet
```{r}
library(graphics)
anscombe <- read.table("Data/anscombe.txt", header=TRUE)
anscombe
summary(anscombe)
head(anscombe)
means <- (c(mean(anscombe$x1), mean(anscombe$x2), mean(anscombe$x3), mean(anscombe$x4)))
means

lm1 <- lm(y1~x1, data=anscombe)
summary(lm1)
lm2 <- lm(y2~x2, data=anscombe)
summary(lm2)
lm3 <- lm(y3~x3, data=anscombe)
summary(lm3)
lm4 <- lm(y4~x4, data=anscombe)
summary(lm4)

# 4 figures arranged in 2 rows and 2 columns
par(mfrow=c(2,2)) #par sets graphical parameters; mfrow defines a vector of how subsequent figures will be organized.
plot(anscombe$x1,anscombe$y1, main="1") 
abline(lm1)
plot(anscombe$x2,anscombe$y2, main="2") 
abline(lm2)
plot(anscombe$x3,anscombe$y3, main="3") 
abline(lm3)
plot(anscombe$x4,anscombe$y4, main="4") 
abline(lm4)
```

